# The Movie Lens Data (Example) {#movie}

## Background

MovieLens (<http://grouplens.org/datasets/movielens/>) is a movie recommendation
engine devel- oped and administered by a research group at the University of
Minnesota. It is very much like the ones included in Netflix and elsewhere, but
free and open. The latest release of the data includes 20,000,000 individual
ratings of 27,000 movies by 138,000 users.

For class, we are using (most of) the much smaller, 1998 release, which includes
100,000 ratings, of 1700 movies by nearly 1000 users. The data included on the
class website is modified slightly from the original. Specifically, we have:

-   added informative column names, which were not included in the original
    data files, but described in a separate readme file
-   removed a couple of columns, including a timestamp indicating when the
    rating was made
-   changed | delimited files to tab delimited,
-   removed 2 movies of unknown genre,
-   removed duplicate entries for a handful of movies that are included in
    the database twice.

It should be noted that some movie names are duplicated, when two different
films happen to have the same name, or when a movie is remade at a later date.
Name and year together make a unique identifier in these cases.

There are 4 data files in the course folder:

-   The ratings themselves are found in **ratingsData.txt**, one rating per
    line. Users and movies are labeled by random, numerical IDs in the
    ratings data.
-   A separate **movieData.txt** file links the numerical movieID to title,
    year, and genre.
-   The **userData.txt** file provides basic demographics for each user,
    including age, gender, occupation and 1zipcode.
-   A fourth file **ratingsDataPlus.txt** combining limited information from
    all 3 files is provided as well for your convenience.

## Goal

To break the ratings for each movie down by user characteristics.


## Data

The first step will be to read in the data.
```{r readData}
userDat    <- read.table("Data/userData.txt",
                         header = T, row.names = 1, as.is = T,
                         sep = "\t", fill = T)
movieDat   <- read.table("Data/movieData.txt",
                         header = T, row.names = 1, as.is = T,
                         sep = "\t", quote = "")
ratingsDat <- read.table("Data/ratingsData.txt",
                         header = T, as.is = T, sep = "\t")
```

After reading all 3 files into R, you will want to think about how you organize
and store the data for analysis. Do your data objects all have useful row and
column names so that you can easily map movies and users across all data sets?
Do you like having the ratings data in a 3 column format, would a matrix make
more sense? Are you happy having everything in 3 separate files? The format used
in the ratingsDataPlus.txt file, may make it easier to work with the data for
some problems. There are trade-offs no matter how its done, though, combining
the files may make it easier to write code for analyses, but it uses a lot of
storage space - the ratingsPlus file alone is more than 3 times as large as the
other 3 files combined.

Logically, data organization comes before analysis, but its not always possible
to anticipate your needs in advance, so you are likely to add things to this
section as you go.

```{r organize}
## make a matrix of ratings
## maybe it will be helpful to have a simple list of movie ids
mov <- rownames(movieDat)
## and a simple list of user ids
usr <- rownames(userDat)

## start with all na's in the matrix
ratMat <- matrix(rep(NA, length(mov) * length(usr)),
                 nrow=length(mov)) #
## add row and col names
rownames(ratMat) <- mov
colnames(ratMat) <- usr

## and fill it up with ratings
for (i in 1:dim(ratingsDat)[1]) {
    n_r <- as.character(ratingsDat[i,2])
    n_c <- as.character(ratingsDat[i,1])
    ratMat[n_r, n_c] <- ratingsDat[i,3]
}


## here is a list of all possible genre names
genres <- colnames(movieDat)[7:20]
```

The age data is numerical, but it might be useful to divide it into groups for
some applications. First I looked at a summary of ages

```{r age}
summary(userDat$age)
```

It is decided that <25, 25-35, 35 - 45 and 45-75 would produce reasonably large
age groups.
```{r age2}

ageGp  <- rep(75, dim(userDat)[1])
ageGp[userDat$age<=45]  <-  45
ageGp[userDat$age<=35]  <-  35
ageGp[userDat$age<=25]  <-  25

summary(ageGp)
```


## Analysis

There are lots of different things we can do here. Here is a function that takes
a movie name as argument and plots age on the x-axis vs ratings on the y-axis.
Color is used to specify gender.

```{r demographics}

### put in the movieID  as argument
ratingsPlot <- function(movID,
                        genderCol = c("M" = "yellow", "F" = "blue")) {
    ## get the real movie name
    name   <- movieDat[movID,"name"]
    ## get the list of users who rated that movie
    raters <- colnames(ratMat)[!is.na(ratMat[movID,])]

    plot(userDat[raters,"age"],  ### need ages for those users
         ratMat[movID,raters],  ### need ratings for those users
         pch = 16,
         col = genderCol[userDat[raters,"gender"]], ### need gender
         main = name, xlab = "age", ylab = "rating")
}

ratingsPlot(rownames(ratMat)[1])
```

```{r demographics2}
ratingsPlot(mov[2])
```

## Results and summary

Most viewers of *Golden Eye* were males, whereas gender distribution of *Toy
Story* was more balanced between males and females.

The viewers of *Toy Story* were younger in generate compared to "Golden Eye".

There is not a clear trend between ratings and age observed in neither of the
two movies.
